---
title: "Pernalonga 2 - Data Prep"
output: html_notebook
---

The first block of code will perform data cleaning based on the data anamolies we found in the first Project. 

```{r}

#First, load all required packages

library(data.table)
library(dplyr)
library(stringr)
library(tidyverse)
library(cluster)
library(factoextra)
library(psych)

#Load the datasets

transactions <- fread("C:/Users/harsh/Desktop/MSBA/Spring/Marketing Analytics/Projects/Project 1/Pernalonga/transaction_table.csv")
products <-fread("C:/Users/harsh/Desktop/MSBA/Spring/Marketing Analytics/Projects/Project 1/Pernalonga/product_table.csv")

##Data Prep

#Map product id's to other product features by joining the two tables
merged_data <- merge(transactions, products, by='prod_id', all=TRUE)

#Handle data anamolies
length(unique(merged_data$tran_id)) #Only 753 unique transaction_id's - Need to make a new transaction id, as the one available is not unique, duplicated very very often and thus prone to errors

#Since each transaction may have multiple rows (depending on number of products types in a transaction), not possible to accurately determinew which transactions were made together
#Way around: Combine customer_id's, store_id's, and purchase date to make a single transaction id as most customers would only make 1 purchase a day at a particular store

#Bring transaction date to date format
merged_data$date <- as.Date(merged_data$tran_dt)

#If a customer has >1 transactions on the same day in the same store, assume it was the same transaction
merged_data$new_transaction_id <- paste(merged_data$cust_id,merged_data$tran_dt,merged_data$store_id)

#Verify the amount paid is consistent with (unit price * quantity - discount amount)
merged_data$transaction_amount <- (merged_data$prod_unit_price * merged_data$tran_prod_sale_qty) - abs(merged_data$tran_prod_discount_amt)

#About 33% product prices are not consistent (10 mil out of 29.6mil)
merged_data$IS_CONSISTENT_FINAL_AMOUNT <- ifelse(merged_data$tran_prod_paid_amt==merged_data$transaction_amount,1,0)
inconsistent_prices <- filter(merged_data,merged_data$IS_CONSISTENT_FINAL_AMOUNT==0) 

#How much is this inconsistency?
head(sort(merged_data$difference,decreasing=T),5) #Very minor differences (e-08 at most), likely due to rounding

#We noticed some transaction amounts are negative, so we removed these from our analysis

merged_data <- merged_data[merged_data$tran_prod_paid_amt>0,]

```

As a first step into EDA, see how many customers bought a private label toothpaste atleast once and then remove from from our data. Our focus will only be on customers who aren't private label toothpaste customers

```{r}

# A restriction for all campaigns except for #11 is customers who purchase Pernalonga's PRIVATE LABEL equivalent of the promoted brand should not be targeted.
# remove transactions that are for PRIVATE LABEL
private_label <- merged_data[merged_data$brand_desc == 'PRIVATE LABEL' & merged_data$category_desc_eng == 'TOOTHPASTE',]
private_label_cust <- data.frame(unique(private_label$cust_id))

# only keep transactions for customers that haven't bought private label
no_private_label <- merged_data[!merged_data$cust_id %in% private_label_cust$unique.private_label.cust_id.,]

#How many customers bought toothpaste that was not private label?
length(unique(no_private_label$cust_id))


```

Next, find loyal customers by doing some calculations. 

1) See each customer's spend on each brand (to look at Colgate as a brand)
2) See each customer's spend on each product category (to look at overall toothpaste sales to each customer)
3) See each customer's spend on each product category from each brand (to look at colgate toothpaste sales for each customer)

```{r}

#Identify each customer's spend on each brand (used for spend on all colgate products)
no_private_label <- no_private_label %>%
  group_by(cust_id,brand_desc) %>%
  mutate(brand_spend= sum(tran_prod_paid_amt))

#Identify each customer's spend on each product category (used for overall spend on Toothpastes (any brand))
no_private_label <- no_private_label %>%
  group_by(cust_id,category_desc_eng) %>%
  mutate(category_spend= sum(tran_prod_paid_amt))


#Identify each customer's spend on each category in brand (used for spend on Toothpaste of each brand)
no_private_label <- no_private_label %>%
  group_by(cust_id,brand_desc,category_desc_eng)%>%
  mutate(brand_category_spend=sum(tran_prod_paid_amt))

#Get transactions limited to toothpaste
toothpaste <- filter(no_private_label,category_desc_eng=="TOOTHPASTE")

#Get transactions for all colgate products
colgate <- filter(no_private_label,brand_desc=="COLGATE")

#Focus on customers and their transactions on either COLGATE products(toothpaste or otherwise) or TOOTHPASTE products (any brand), accounting for overlapping transactions only once
colgate_or_anytoothpaste <- unique(rbind(toothpaste,colgate))

colgate_or_anytoothpaste <- distinct_at(colgate_or_anytoothpaste,.vars = c("cust_id","brand_desc","category_desc_eng") ,.keep_all = T)

#For each customer, obtain amount spent on colgate products

colgate_spend <- filter(colgate_or_anytoothpaste,brand_desc=="COLGATE")
colgate_spend <- colgate_spend[,c("cust_id","brand_spend")]
colgate_spend <- distinct_at(colgate_spend,.vars="cust_id",.keep_all = T)

#For each customer, obtain amount spent on colgate toothpaste only

colgate_toothpaste_spend <- filter(colgate_or_anytoothpaste,brand_desc=="COLGATE" & category_desc_eng=="TOOTHPASTE")
colgate_toothpaste_spend <- colgate_toothpaste_spend[,c("cust_id","brand_category_spend")]
colgate_toothpaste_spend <- distinct_at(colgate_toothpaste_spend,.vars="cust_id",.keep_all = T)

#For each customer, obtain amount spent on all toothpastes
toothpaste_spend <- filter(colgate_or_anytoothpaste,category_desc_eng=="TOOTHPASTE")
toothpaste_spend <- toothpaste_spend[,c("cust_id","category_spend")]
toothpaste_spend <- distinct_at(toothpaste_spend,.vars="cust_id",.keep_all = T)

#Get list of these customers and store their total spends on toothpaste, colgate, and colgate toothpaste
target_customers <- as.data.frame(unique(colgate_or_anytoothpaste$cust_id))
names(target_customers) <- "cust_id"
target_customers <- plyr::join_all(list(target_customers,colgate_spend,colgate_toothpaste_spend,toothpaste_spend),by="cust_id",type="left")

#Replace NA's with 0 as some customers did not purchase in all 3 categories

target_customers[is.na(target_customers)] <- 0

#Rename columns for convenience and easy understanding
names(target_customers) <- c("cust_id","Colgate_Spend","Colgate_Toothpaste_Spend","Total_Toothpaste_Spend")

```

We have a list of 3596 target customers - let's split them into "baseline segment", and "target for promotion segments"

Criteria for baseline segment: Loyal colgate toothpaste customer who
a) Spends atleast 70% of his money spent on toothpaste on Colgate 
b) Is in the top quartile of colgate toothpaste customers based on overall sales

```{r}

#Define cutoffs to be selected as baseline
colgate_toothpaste_spend_cutoff <- quantile(target_customers$Colgate_Toothpaste_Spend,0.75)
colgate_toothpaste_ratio_cutoff <- 0.7

#Calculate ratio of colgate toothpaste spend to total toothpaste spend
target_customers$toothpaste_ratio <- target_customers$Colgate_Toothpaste_Spend/target_customers$Total_Toothpaste_Spend

#Identify baseline customers
baseline_customers <- filter(target_customers,target_customers$Colgate_Toothpaste_Spend >= colgate_toothpaste_spend_cutoff & target_customers$toothpaste_ratio>=colgate_toothpaste_ratio_cutoff)

#Customers in the target list that are not in the baseline category
remaining_customers <- setdiff(target_customers,baseline_customers)

#Next, let's setup a list of customers we will send personalized promotions to based on similarity with baseline customers

#As we recommend launching a pilot, we only want to target customers that a) Spend on colgate products (other than toothpaste), or b) Spend on other toothpastes

#Discard customers with very low spends on both (25th percentile and below) because they likely wont purchase Colgate toothpaste based on their historical sales

promotion_customers <- filter(remaining_customers,remaining_customers$Colgate_Spend> quantile(remaining_customers$Colgate_Spend,0.25) | remaining_customers$Total_Toothpaste_Spend >quantile(remaining_customers$Total_Toothpaste_Spend,0.25))

promotion_customer_ids <- promotion_customers$cust_id

```

Find the products that these baseline customers spent the most money on. We will find similar users based on their spend on these top products

```{r}

#Store customer id's for these baseline customers
baseline_customer_ids <- unique(baseline_customers$cust_id)

#Calculate the money spent by each baseline customer on each product 

no_private_label <- no_private_label %>%
  group_by(cust_id,prod_id) %>%
  mutate(prod_category_spend= sum(tran_prod_paid_amt))

#Keep only the baseline customers and keep relevant columns only
baseline_no_private_label <- no_private_label[no_private_label$cust_id %in% baseline_customer_ids, c("cust_id","prod_id","tran_prod_paid_amt","prod_category_spend")]

#Keep one row per customer per product category
baseline_no_private_label <- distinct_at(baseline_no_private_label,.vars=c("cust_id","prod_id"),.keep_all = T)

#Calculate overall spend on each product from all baseline customers in total -- This will be used to shortlist the Top X products with Maximum money spent

baseline_no_private_label <- baseline_no_private_label %>%
  group_by(prod_id) %>%
  mutate(total_prod_sales= sum(tran_prod_paid_amt))

#Keep relevant columns and keep only 1 row per product 
prod_sales_baseline <- distinct_at(baseline_no_private_label,.vars = "prod_id",.keep_all = T)
prod_sales_baseline <- prod_sales_baseline[,c("prod_id","total_prod_sales")]

#Sort by descending order of spend
prod_sales_baseline <- arrange(prod_sales_baseline,desc(total_prod_sales))

#How many products to get 80% sales? We are not considering the bottom 20% products for this segment because it would bring noise to the data and increase computation time

cumsum(prod_sales_baseline$total_prod_sales)[5000]/sum(prod_sales_baseline$total_prod_sales) #82%

#Identify top X products for baseline customers
top_n <- 5000
top_products <- prod_sales_baseline$prod_id[1:top_n] #Stores product id's for top X products


```

Now, make a dataframe that contains all baseline customers (rows) and their individual spends on these top 5000 products

```{r}

#Keep only observations for the top 5000 products for baseline customers
baseline_no_private_label <- baseline_no_private_label[baseline_no_private_label$prod_id %in% top_products, c("cust_id","prod_id","prod_category_spend")]

#Change data from long to wide format - customers in rows, products in columns, product spend as values
baseline_no_private_label <- pivot_wider(baseline_no_private_label,names_from=prod_id,values_from=prod_category_spend)

#Replace NA's with 0s
baseline_no_private_label[is.na(baseline_no_private_label)] <- 0

```

Finally, make a dataframe that contains all targeted customers (rows) and their individual spends on these top 5000 products (same products as baseline customers)

```{r}

target_no_private_label <- no_private_label[no_private_label$cust_id %in% promotion_customer_ids, c("cust_id","prod_id","prod_category_spend")]

#Keep one row per customer per product category
target_no_private_label <- distinct_at(target_no_private_label,.vars=c("cust_id","prod_id"),.keep_all = T)

#Keep only observations for the top X products that similarity will be measured on
target_no_private_label <- target_no_private_label[target_no_private_label$prod_id %in% top_products,]

#Change data from long to wide format - customers in rows, products in columns, product spend as values
target_no_private_label <- pivot_wider(target_no_private_label,names_from=prod_id,values_from=prod_category_spend)

#Replace NA's with 0s
target_no_private_label[is.na(target_no_private_label)] <- 0


```

Save the files to measure similarity later - Performed in "Pernalonga_PSIM_and_NaiveBayes.R"

```{r}

fwrite(target_no_private_label,"Target_customers_input.csv")
fwrite(baseline_no_private_label,"baseline_customers_input.csv")
fwrite(no_private_label,"no_private_label.csv")

```


Naive Bayes data prep below


```{r}

#Calculate Probability for colgate toothpaste purchase on the target universe (including those we discarded based on similarity)

num_customers <- length(unique(no_private_label$cust_id))
colgate_toothpaste_transactions <- filter(no_private_label,category_desc_eng=="TOOTHPASTE" & brand_desc=="COLGATE")
colgate_customers <- unique(colgate_toothpaste_transactions$cust_id)
num_colgate_toothpaste_cust <- length(unique(colgate_toothpaste_transactions$cust_id))

#Calculate probabilities
PBuyColgate <- num_colgate_toothpaste_cust/num_customers
PNotBuyColgate <- 1 - PYes

```

Now calculate mean and standard deviation for each product across the customer base (3799 customers)

```{r}

#Spread the dataframe - customers in rows, 5k products in columns and overall sales as values
universe_no_private_label <- no_private_label[no_private_label$prod_id %in% top_products,c("cust_id","prod_id","prod_category_spend")]
universe_no_private_label <- distinct_at(universe_no_private_label,.vars = c("cust_id","prod_id") ,.keep_all = T)
universe_no_private_label <- pivot_wider(universe_no_private_label,names_from=prod_id,values_from=prod_category_spend)
universe_no_private_label[is.na(universe_no_private_label)] <- 0

#Colgate customers only

colgate_customers_spendbyproduct <- universe_no_private_label[universe_no_private_label$cust_id %in% colgate_customers,]
non_colgate_customers_spendbyproduct <- universe_no_private_label[!universe_no_private_label$cust_id %in% colgate_customers,]

#Calculate mean and standard deviation for these two customer groups

colgate_cust_meanspend <- colMeans(colgate_customers_spendbyproduct[,2:ncol(colgate_customers_spendbyproduct)])
colgate_cust_sdspend <- apply(colgate_customers_spendbyproduct[,2:ncol(colgate_customers_spendbyproduct)],2,sd)

non_colgate_cust_meanspend <- colMeans(non_colgate_customers_spendbyproduct[,2:ncol(non_colgate_customers_spendbyproduct)])
non_colgate_cust_sdspend <- apply(non_colgate_customers_spendbyproduct[,2:ncol(non_colgate_customers_spendbyproduct)],2,sd)

#Save these two as dataframes

colgate_cust_summary <- as.data.frame(rbind(colgate_cust_meanspend,colgate_cust_sdspend))
non_colgate_cust_summary <- as.data.frame(rbind(non_colgate_cust_meanspend,non_colgate_cust_sdspend))

fwrite(colgate_cust_summary,"colgate_cust_summary.csv")
fwrite(non_colgate_cust_summary,"non_colgate_cust_summary.csv")

#Did each customer buy colgate toothpaste in the past?
colgate_customers_transactions$Bought_Colgate <- ifelse(colgate_customers_transactions$cust_id %in% colgate_customers,1,0)

#Save each target customer's sales data and whether they bought colgate toothpaste or not
fwrite(colgate_customers_transactions,"all_customers_sales.csv") #file used in "PSIM_and_NaiveBayes.R"


```

Finally, complete Incremental Sales and Volume analysis based on target customers & loyal customers output from "PSIM_and_NaiveBayes.R"

```{r}

#Load dataframe of each customer to be targetted, their segment & probability of purchase
promo_list <-fread("C:/Users/harsh/Desktop/MSBA/Spring/Marketing Analytics/Projects/Project 2/promo_list.csv")


#Make mode function
getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}

#Add segment to each customer id (baseline vs target)
colgate_toothpaste_only <- left_join(colgate_toothpaste_only,promo_list[,c("cust_id","cust_type")])

#Calculate segment level stats for ROI analysis - mean paid amount, sale price without discount, number of units sold (for historical distribution)
colgate_tooth_unitprice_by_subcat <- colgate_toothpaste_only %>%
  group_by(sub_category_desc,cust_type) %>%
  mutate(most_common_paid_price= mean(tran_prod_paid_amt/tran_prod_sale_qty),
         most_common_sale_price=getmode(tran_prod_sale_amt/tran_prod_sale_qty),
         units_sold = sum(tran_prod_sale_qty))

#Calculate current sales distribution between subcategories
colgate_tooth_unitprice_by_subcat <- colgate_tooth_unitprice_by_subcat %>%
  group_by(cust_type) %>%
  mutate(distribution= units_sold/sum(tran_prod_sale_qty))

#Keep 1 row for each sub-category to get all statistics on the sub-category level
colgate_tooth_unitprice_by_subcat <-  distinct_at(colgate_tooth_unitprice_by_subcat[,c("sub_category_desc","most_common_paid_price","cust_type","most_common_sale_price","distribution")],.vars = "sub_category_desc",.keep_all = T)

total_expected_sales_baseline <- 696 #Calculate based on Customers targeted * Avg units/transaction in that segment
total_expected_sales_target <- 309 

#Expected units sold = Units distribution for that sub-category in that segment * Overall units expected to be sold in that segment
colgate_tooth_unitprice_by_subcat$Expected_Units_sold <- ifelse(colgate_tooth_unitprice_by_subcat$cust_type=="Baseline",colgate_tooth_unitprice_by_subcat$distribution*total_expected_sales_baseline,colgate_tooth_unitprice_by_subcat$distribution*total_expected_sales_target)

fwrite(colgate_tooth_unitprice_by_subcat,"ROI.csv") #Output for report

```

